---
title: "Reproducible Research: Peer Assessment 1"
author: "Dana Jacob"
date: "July 11, 2015"
output: html_document
---

library(plyr)

####Loading and preprocessing the data

```{r read_data}
unzip("activity.zip")
rawData <- read.csv("activity.csv", header = TRUE, sep = ",")
rawData$date <- as.Date(rawData$date)    
```
    
####What is mean total number of steps taken per day?

#####First, we calculate total number of steps each day and plot a histogram

```{r total_steps_histogram}
stepsDay <- tapply(rawData$steps, rawData$date, FUN=sum, na.rm=TRUE)
hist(stepsDay, breaks=20, main = "Total number of steps per day",
     xlab = "Number of steps per day", ylab = "Number of days")
```

#####Next, we calculate mean and median number of steps per day

```{r mean_median}
meanDay <- mean(stepsDay, na.rm = TRUE)
round(meanDay, digits = 0)
medianDay <- median(stepsDay, na.rm = TRUE)
round(medianDay, digits = 0)
```

Average number of steps per day = 9,354  
Median number of steps per day = 10,395

####What is the average daily activity pattern?

#####We plot a time series of average number of steps per 5-minute interval across all days

```{r time_series}
library(ggplot2)

meanInterval <- aggregate(x = list(steps=rawData$steps),
                by = list(interval=rawData$interval), 
                FUN = mean, na.rm = TRUE)

ggplot(meanInterval, aes(x=interval, y=steps)) + 
    geom_line() +
    ggtitle("Average number of steps per interval") +
    xlab("5-minute intervals") +
    ylab("Average number of steps")
```

#####We identify the 5-minute interval that has the maximum number of steps

```{r maximum_steps}
maxStep <- meanInterval[which.max(meanInterval$steps),]
round(maxStep, digits = 0)
```

At interval 835, which occured on Oct 1, 2012, 206 steps were taken, higher than number of steps taken at any other intervals. 

####Imputing missing values

#####First, we calculate and report the total number of missing values in the dataset 

```{r number_missing}
table (is.na(rawData))
```

There are 2,304 missing values in the data. 

#####Next, we use the mean for the 5-minute intervals to impute missing values and create a new dataset

```{r new_data_fill_missing}
newData <- rawData
newData$steps[is.na(newData$steps)] <- 
    tapply(newData$steps, newData$interval, mean, na.rm = TRUE)

#Check to make sure missing values are filled
table(is.na(newData))
```

####Are there differences in activity patterns between weekdays and weekends?

#####1. We plot a histogram of total number of steps per day using the new dataset

```{r new_total_steps_histogram}
stepsDayNew <- tapply(newData$steps, newData$date, FUN=sum, na.rm=TRUE)
hist(stepsDayNew, breaks=20, main = "Total number of steps per day using new dataset",
     xlab = "Number of steps per day", ylab = "Number of days")
```

The histogram using the new dataset is closer to a 'normal distribution' as filled-in values for missing days helped to 'normalize' the data.

#####2. We calculate mean and median number of steps per day using the new dataset

```{r new_mean_median}
meanDayNew <- mean(stepsDayNew, na.rm = TRUE)
round(meanDayNew, digits = 0)
medianDayNew <- median(stepsDayNew, na.rm = TRUE)
round(medianDayNew, digits = 0)
```

Once missing values are filled in with interval means, the average and median number of steps per day increased:  
Average number of steps per day - old=9,354 new=10,766  
Median number of steps per day - old=10,395 new=10,766

#####3. We add a factor variable, dayType, to the new dataset to indicate whether a given date is a weekday or weekend day

```{r day_type_factor}
newData$date <- as.Date(newData$date)
newData$dayType <- factor(format(newData$date, "%A"))
levels(newData$dayType) <- 
    list(weekday=c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"),
         weekend=c("Saturday", "Sunday"))
```

#####4. We make a panel plot with time series to compare activity patterns between weekdays and weekends

```{r panel_plot}
stepsIntervalNew <- aggregate(steps ~ interval + dayType,
                             data = newData, FUN = mean)

library(lattice)
xyplot(type = "l", data = stepsIntervalNew,
       steps ~ interval | dayType,
       layout = c(1,2), grid = T,
       xlab ="Interval",
       ylab = "Number of steps")
```

Conclusion: more steps were taken on weekends than on weekdays.
